---
title: "Small Molecules"
author: "Brian S. Yandell"
date: "8/29/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goal: Set up small molecules data for qtl2shiny and other uses.

See Box folder [MS data sets for DO project](https://uwmadison.app.box.com/folder/33549545179)
and subfolders

* RDS files for all QTLs (genome scans)
* Gatti R scripts
* Raw and UW normalized data files

See also [Dan Gatti's AttieMetabolomics GitHub](https://github.com/dmgatti/AttieMetabolomics) and
[Dan Gatti's FTP site](ftp://ftp.jax.org/dgatti/_for_Mark/liver/).

Types of molecular measurements:

* liver lipids
* cecum lipids
* plasma lipids
* liver metabolites

There may be others coming.

Here are the relevant files:

size | type | name
---- | -------------- | ------------
961K | mouse by trait | attie_liver_lipids_zscore_normalized.rds
225K | mouse by trait | attie_liver_metabolites_zscore_normalized.rds
1.2M | mouse by trait | attie_plasma_lipids_zscore_normalized.rds
1.6M | mouse by trait | attie_cecum_lipids_zscore_normalized.rds
176K | peaks | cecum_lipids_jax_norm_qtl_summary_thresh_6.csv
132K | peaks | liver_lipids_jax_norm_qtl_summary_thresh_6.csv
20K | peaks | liver_metabolites_jax_norm_qtl_summary_thresh_6.csv
193K | peaks | plasma_lipids_jax_norm_qtl_summary_thresh_6.csv

The files `*_pheno.descr.rds` are relevant only in identifying the covariates used.
They are different from what Karl used as dictionary. 

```{r message=FALSE}
library(stringr)
library(dplyr)
library(tidyr)
```

```{r}
datapath <- "~/Documents/Research/attie_alan/DO/data/DerivedData/"
```

Phenotypes follow covariates in the mouse by trait tables, and the `Mouse.ID` is `DO-nnn`, as apposed to `nn` and `nnn` in Karl's work. All fixable but will take careful examination. 

Dan used `generation` as covariate, but there are errors in this column. Instead,
I will use `DOwave`.

```{r}
liverz <- readRDS(file.path(datapath, "molecule", 
                            "attie_liver_lipids_zscore_normalized.rds"))
metabz <- readRDS(file.path(datapath, "molecule",
                            "attie_liver_metabolites_zscore_normalized.rds"))
plasmaz <- readRDS(file.path(datapath, "molecule",
                            "attie_plasma_lipids_zscore_normalized.rds"))
cecumz <- readRDS(file.path(datapath, "molecule",
                            "attie_cecum_lipids_zscore_normalized.rds"))
```

## Pull out relevant phenotypes and covariates and analyses stuff.

We need the following to tranform into Karl type files:

* `analyses_tbl`: `c("pheno","output","pheno_group","longname","pheno_type","model","transf","offset","winsorize",
"sex","diet_days","Ins_per_islet","Glu_0min","Ins_0min","weight_sac","oGTT_weight",
"DOwave2","DOwave3","DOwave4","DOwave5","Gcg_content")`
* `peaks_tbl`: `c("pheno","longname","output","pheno_group","pheno_type","chr","lod","pos")`
* `pheno_tbl`: data frame with mouseID as rownames, pheno as colnames
* `covar_mx`: matrix with mouseID as rownmaes, covar as colnames

Need to modify `covar` from matrix to data frame (and change code in qtl2shiny), and also modify `analyses_tbl`. Need to coordinate change to data frame in particular with use of `DOwave` as covariate.
Need to extract covariates, compare that Mouse.ID, sex, wave, chrM, chrY agree across all sets. Then need to give Batch unique names and combine.
Need to extract phenos and put in form to be reused.

### Peaks

The molecules of the same type (lipids, metabolites) may overlap from one tissue to another. Therefore we prepend with name of tissue.

```{r}
setup_molecule_peaks <- function(datapath, filename, typename, tissue) {
  read.csv(file.path(datapath, "molecule", filename),
           stringsAsFactors = FALSE) %>%
    rename(pheno = analyte) %>%
    mutate(pheno = paste(tissue, pheno, sep = "."),
           longname = pheno,
           output = pheno,
           pheno_type = typename,
           pheno_group = "Molecule") %>%
    select(pheno,longname,output,pheno_group,pheno_type,chr,lod,pos)
}
```

```{r}
peaks_molecule <- bind_rows(
  setup_molecule_peaks(datapath,
                       "liver_lipids_jax_norm_qtl_summary_thresh_6.csv",
                       "LiverLipid",
                       "Liver"),
  setup_molecule_peaks(datapath,
                       "plasma_lipids_jax_norm_qtl_summary_thresh_6.csv",
                       "PlasmaLipid",
                       "Plasma"),
  setup_molecule_peaks(datapath,
                       "liver_metabolites_jax_norm_qtl_summary_thresh_6.csv",
                       "LiverMetab",
                       "Liver"),
  setup_molecule_peaks(datapath,
                       "cecum_lipids_jax_norm_qtl_summary_thresh_6.csv",
                       "CecumLipid",
                       "Cecum")) %>%
  mutate(chr = factor(chr, c(1:19, "X")))
```

```{r}
str(peaks_molecule)
```

```{r}
saveRDS(peaks_molecule, file.path(datapath, "molecule", "peaks_small.rds"))
```

### Analyses

```{r}
analyses_molecule <- peaks_molecule %>%
  distinct(pheno, output, pheno_group, longname, pheno_type) %>%
  mutate(model = "normal",
         transf = "identity",
         offset = 0,
         sex = TRUE,
         DOwave = TRUE,
         BatchLiverLipid = (pheno_type == "LiverLipid"),
         BatchCecumLipid = (pheno_type == "CecumLipid"),
         BatchLiverMetab = (pheno_type == "LiverMetab"),
         BatchPlasmaLipid = (pheno_type == "PlasmaLipid"))
```

```{r}
str(analyses_molecule)
```

```{r}
saveRDS(analyses_molecule, file.path(datapath, "molecule", "analyses_small.rds"))
```

### Covariates

We will add columns for `Batch` (specific to `pheno_type`) and for `chrM` and `chrY`, which might be useful in the future. Also add `coat_color`. All other covariates are ignored here.

```{r}
covar_molecule <- liverz %>% 
  select(Mouse.ID,coat_color,Batch,chrM,chrY) %>%
  mutate(Batch = as.character(Batch)) %>%
  rename(BatchLiverLipid = Batch)
covar_molecule <- left_join(
  covar_molecule,
  plasmaz %>% 
    select(Mouse.ID, Batch) %>%
    mutate(Batch = as.character(Batch)) %>%
    rename(BatchPlasmaLipid = Batch),
  by = "Mouse.ID")
covar_molecule <- left_join(
  covar_molecule,
  metabz %>% 
    select(Mouse.ID, Batch) %>%
    mutate(Batch = as.character(as.numeric(str_replace(Batch, "Batch", "")))) %>%
    rename(BatchLiverMetab = Batch),
  by = "Mouse.ID")
covar_molecule <- left_join(
  covar_molecule,
  cecumz %>% 
    select(Mouse.ID, Batch) %>%
    mutate(Batch = as.character(Batch)) %>%
    rename(BatchCecumLipid = Batch),
  by = "Mouse.ID")
rownames(covar_molecule) <- 
  as.numeric(str_replace(covar_molecule$Mouse.ID, "DO-", ""))
covar_molecule$Mouse.ID <- NULL
```

```{r}
str(covar_molecule)
```

```{r}
saveRDS(covar_molecule, file.path(datapath, "molecule", "covar_small.rds"))
```

## Phenotypes

Phenotypes are in files with names like `attie_liver_lipids_zscore_normalized.rds`. That is, they are Z-scores of Jax-normalized data. These files have phenotypes prepended by covariates and some extra columns for mitochondria and chr `Y` (`chrM`, `chrY`). 

Approach here is to match `Mouse.ID` and pick up phenotypes. These are large files (1Gb).

```{r}
pheno_prepend <- function(object, tissue) {
  object %>%
    select(-(wave:chrY)) %>%
    rename_at(vars(-Mouse.ID), function(x) paste(tissue, x, sep = "."))
}
```

```{r}
pheno_molecule <- full_join(
  liverz %>% pheno_prepend("Liver"),
  plasmaz %>% pheno_prepend("Plasma"),
  by = "Mouse.ID")
pheno_molecule <- full_join(
  pheno_molecule,
  metabz %>% pheno_prepend("Liver"),
  by = "Mouse.ID")
pheno_molecule <- full_join(
  pheno_molecule,
  cecumz %>% pheno_prepend("Cecum"),
  by = "Mouse.ID")
rownames(pheno_molecule) <- 
  as.numeric(str_replace(pheno_molecule$Mouse.ID, "DO-", ""))
pheno_molecule$Mouse.ID <- NULL
```

```{r}
dim(pheno_molecule)
```

```{r}
pheno_molecule[1:5,1:3]
```

```{r}
saveRDS(pheno_molecule, file.path(datapath, "molecule", "pheno_small.rds"))
```
