---
title: "Small Molecules"
author: "Brian S. Yandell"
date: "8/29/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goal: Set up small molecules data for qtl2shiny and other uses.

See Box folder [MS data sets for DO project](https://uwmadison.app.box.com/folder/33549545179)
and subfolders

* RDS files for all QTLs (genome scans)
* Gatti R scripts
* Raw and UW normalized data files

See also [Dan Gatti's AttieMetabolomics GitHub](https://github.com/dmgatti/AttieMetabolomics) and
[Dan Gatti's FTP site](ftp://ftp.jax.org/dgatti/_for_Mark/liver/).

Types of molecular measurements:

* liver lipids
* cecum lipids
* plasma lipids
* liver metabolites

Here are the files I think are relevant:

size | type | name
---- | -------------- | ------------
961K | mouse by trait | attie_liver_lipids_zscore_normalized.rds
225K | mouse by trait | attie_liver_metabolites_zscore_normalized.rds
1.2M | mouse by trait | attie_plasma_lipids_zscore_normalized.rds
1.6M | mouse by trait | attie_cecum_lipids_zscore_normalized.rds
176K | peaks | cecum_lipids_jax_norm_qtl_summary_thresh_6.csv
132K | peaks | liver_lipids_jax_norm_qtl_summary_thresh_6.csv
20K | peaks | liver_metabolites_jax_norm_qtl_summary_thresh_6.csv
193K | peaks | plasma_lipids_jax_norm_qtl_summary_thresh_6.csv
55K | description | cecum_lipid_pheno.descr.rds
31K | description | liver_lipid_pheno.descr.rds
2.2K | description | liver_metabolite_pheno.descr.rds
41K | description | plasma_lipids_pheno.descr.rds
| untransformed | attie_liver_lipids_normalized.rds
| genome scan | liver_lipids_jax_norm_all_qtl.rds

```{r message=FALSE}
library(stringr)
library(dplyr)
library(tidyr)
```

```{r}
datapath <- "~/Documents/Research/attie_alan/DO/data/DerivedData/"
```

The `pheno.descr` files are different from what Karl used as dictionary, so will have to construct that manually. Phenotypes follow covariates in the mouse by trait tables, and the `Mouse.ID` is `DO-nnn`, as apposed to `nn` and `nnn` in Karl's work. All fixable but will take careful examination. 

```{r}
liverz <- readRDS(file.path(datapath, "molecule", 
                            "attie_liver_lipids_zscore_normalized.rds"))
metabz <- readRDS(file.path(datapath, "molecule",
                            "attie_liver_metabolites_zscore_normalized.rds"))
plasmaz <- readRDS(file.path(datapath, "molecule",
                            "attie_plasma_lipids_zscore_normalized.rds"))
cecumz <- readRDS(file.path(datapath, "molecule",
                            "attie_cecum_lipids_zscore_normalized.rds"))
```

#### Check agreement on covariates

```{r}
covar <- readRDS(file.path(datapath, "covar.rds"))
if(is.null(covar$DOwave)) {}
  covar <- cbind(covar, DOwave = 1 + covar[,"DOwave2"] + 2 * covar[,"DOwave3"] +
    3 * covar[,"DOwave4"] + 4 * covar[,"DOwave5"])
}
```

Check sex.

```{r}
sexl <- liverz$sex
names(sexl) <- as.numeric(stringr::str_replace(liverz$Mouse.ID, "DO-", ""))
tmp <- qtl2scan::get_common_ids(covar, sexl)
table(covar[tmp, "sex"], sexl[tmp])
```

check wave and generation

```{r}
wave <- liverz$wave
names(wave) <- as.numeric(str_replace(liverz$Mouse.ID, "DO-", ""))
tmp <- qtl2scan::get_common_ids(covar, wave)
table(covar[tmp, "DOwave"], wave[tmp])
```

```{r}
gen <- liverz$generation
names(gen) <- as.numeric(str_replace(liverz$Mouse.ID, "DO-", ""))
tmp <- qtl2scan::get_common_ids(covar, gen)
table(covar[tmp, "DOwave"], gen[tmp])
```

```{r}
table(covar[,"DOwave"])
table(liverz$generation)
```

```{r}
tmp2 <- covar[tmp, "DOwave"] < 3 & gen[tmp] == 19
tmp2 <- names(tmp2[tmp2])
```

```{r}
covar[tmp2,]
```

```{r}
liverz %>%
  filter(as.numeric(str_replace(Mouse.ID, "DO-", "")) %in% tmp2) %>%
  select(1:11)
```

## Pull out relevant phenotypes and covariates and analyses stuff.

We need the following to tranform into Karl type files:

* `analyses_tbl`: `c("pheno","output","pheno_group","longname","pheno_type","model","transf","offset","winsorize",
"sex","diet_days","Ins_per_islet","Glu_0min","Ins_0min","weight_sac","oGTT_weight",
"DOwave2","DOwave3","DOwave4","DOwave5","Gcg_content")`
* `peaks_tbl`: `c("pheno","longname","output","pheno_group","pheno_type","chr","lod","pos")`
* `pheno_tbl`: data frame with mouseID as rownames, pheno as colnames
* `covar_mx`: matrix with mouseID as rownmaes, covar as colnames

Details to be added here. Need to modify `covar`, save as data frame, and also modify `analyses_tbl`. Need to coordinate change to data frame in particular with use of `DOwave` as covariate.

Need to extract covariates, compare that Mouse.ID, sex, wave, chrM, chrY agree across all sets. Then need to give Batch unique names and combine.

Need to extract phenos and put in form to be reused.

Need to get analyses info right.

Need to go to CSV files to get peaks.

### Phenotypes

Phenotypes are in files with names like `attie_liver_lipids_zscore_normalized.rds`. That is, they are Z-scores of Jax-normalized data. These files have phenotypes prepended by covariates and some extra columns for mitochondria and chr `Y` (`chrM`, `chrY`). 

Approach here is to match `Mouse.ID` and pick up phenotypes. These are large files (1Gb).

### Peaks

* `peaks_tbl`: `c("pheno","longname","output","pheno_group","pheno_type","chr","lod","pos")`

These have names like `liver_lipids_jax_norm_qtl_summary_thresh_6.csv`.

```{r}
setup_molecule_peaks <- function(datapath, filename, groupname) {
  read.csv(file.path(datapath, "molecule", filename),
           stringsAsFactors = FALSE) %>%
    rename(pheno = analyte) %>%
    mutate(longname = pheno,
           output = pheno,
           pheno_type = "molecule",
           pheno_group = groupname) %>%
    select(pheno,longname,output,pheno_group,pheno_type,chr,lod,pos)
}
```

```{r}
peaks_liver_lipid <- 
  setup_molecule_peaks(datapath,
                       "liver_lipids_jax_norm_qtl_summary_thresh_6.csv",
                       "liver_lipid")
peaks_plasma_lipid <- 
  setup_molecule_peaks(datapath,
                       "plasma_lipids_jax_norm_qtl_summary_thresh_6.csv",
                       "plasma_lipid")
peaks_liver_metab <- 
  setup_molecule_peaks(datapath,
                       "liver_metabolites_jax_norm_qtl_summary_thresh_6.csv",
                       "liver_metab")
peaks_cecum_lipid <- 
  setup_molecule_peaks(datapath,
                       "cecum_lipids_jax_norm_qtl_summary_thresh_6.csv",
                       "cecum_lipid")
```

Do we have to adjust names that overlap? Yes. Need to adjust here and in the `pheno` data frame.

```{r}
llip <- sort(unique(peaks_liver_lipid$pheno))
plip <- sort(unique(peaks_plasma_lipid$pheno))
clip <- sort(unique(peaks_cecum_lipid$pheno))
c(liver = length(llip), plasma = length(plip), cecum = length(clip))
table(table(c(llip,plip)))
table(table(c(llip,clip)))
table(table(c(clip,plip)))
table(table(c(llip,clip,plip)))
tmp <- table(c(llip,clip,plip))
names(tmp[tmp == 2])
```

Now pull out unique `pheno` and make `analyses_molecule`.

* `analyses_tbl`: `c("pheno","output","pheno_group","longname","pheno_type","model","transf","offset","winsorize",
"sex","diet_days","Ins_per_islet","Glu_0min","Ins_0min","weight_sac","oGTT_weight",
"DOwave2","DOwave3","DOwave4","DOwave5","Gcg_content")`

### Covariates matrix and Analyses table

Strategy is to make `covar` a data frame (convert from current format as numeric matrix)
and use `model.matrix`. We will add columns for `Batch` (note names above) and for `chrM` and `chrY`, which might be useful in the future. Also add `coat_color`. All other covariates are ignored here.

Here, covariates are indicated in `liverd$covar_list` as colon-separated names, but all are `sex:generation:Batch`. 
`Batch` is distinct (I think) for each type of molecule.
`generation` is DO generation, which should agree with `DOwave` (1=17, 2=18, 3=19, 4=21, 5=2_).
Turns out `generation` has 5 discrepancies. We will use `DOwave`

The covariate values are with phenotypes as first several columns and some extra columns for mitochondria and chr `Y` (`chrM`, `chrY`).

```{r}
liverz %>%
  count(chrM, chrY) %>%
  spread(chrM, n)
```

#### Check if all equal across small molecules

Comparisons are made below between liver lipids and the other three analytes.

```{r}
all.equal(liverz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY),
          plasmaz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY))
```

```{r}
table(liverz$Batch,plasmaz$Batch)
```

```{r}
all.equal(liverz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY),
          metabz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY))
```

```{r}
matchM <- qtl2scan::get_common_ids(liverz$Mouse.ID, metabz$Mouse.ID)
matchM2 <- which(diff(match(metabz$Mouse.ID, matchM)) != 1)
matchM2 <- sort(unique(c(matchM2 - 1, matchM2, matchM2 + 1)))
tmp <- rbind(metabz$Mouse.ID[matchM2], liverz$Mouse.ID[matchM2])
dimnames(tmp) <- list(c("metab","liver"), matchM2)
tmp
```


```{r}
matchM2 <- match(matchM, metabz$Mouse.ID)
all.equal(liverz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY),
          metabz[matchM2,] %>% select(Mouse.ID,coat_color,Batch,chrM,chrY))
```

```{r}
table(liverz$Batch,as.numeric(str_replace(metabz[matchM2,"Batch"], "Batch", "")))
```

Liver and metab differ in `Batch` for `r sum(!(liverz$Batch == as.numeric(str_replace(metabz[matchM2,"Batch"], "Batch", ""))))` mice.
This is a different mouse: `r liverz$Mouse.ID[which(!(liverz$Batch == as.numeric(str_replace(metabz[matchM2,"Batch"], "Batch", ""))))]`.

Cecum has 3 fewer mice than the other assays. Batch is substantially different for cecum.

```{r}
all.equal(liverz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY),
          cecumz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY))
```

```{r}
matchC <- qtl2scan::get_common_ids(liverz$Mouse.ID, cecumz$Mouse.ID)
matchC2 <- match(matchC, liverz$Mouse.ID)
all.equal(liverz[matchC2,] %>% select(Mouse.ID,coat_color,Batch,chrM,chrY),
          cecumz %>% select(Mouse.ID,coat_color,Batch,chrM,chrY))
```

```{r}
table(liverz[matchC2,"Batch"],cecumz$Batch)
```

Liver and cecum differ in `Batch` for `r sum(!(liverz[matchC2,"Batch"] == cecumz$Batch))` mice.

Conclusions:

* Liver and plasma have different `Batch` for `r sum(!(liverz$Batch == plasmaz$Batch))` mice.
* Metab has `Mouse.ID` `DO-191` in the wrong place. `Batch` is coded as `Batchnn` rather than `nn`,
and differs for mouse `r liverz$Mouse.ID[which(!(liverz$Batch == as.numeric(str_replace(metabz[matchM2,"Batch"], "Batch", ""))))]`.
* Cecum has 3 fewer mice than the other assays. Cecum `Batch` differs for `r sum(!(liverz[matchC2,"Batch"] == cecumz$Batch))` mice.

Need to add `generation` and `Batch`, but there will be four `Batch` (at least):
`Batch_liver_lipid`, `Batch_plasma_lipids`, `Batch_cecum_lipids`, `Batch_liver_metabolites`.

```{r}
covar_small <- liverz %>% 
  select(Mouse.ID,coat_color,Batch,chrM,chrY) %>%
  mutate(Batch = as.character(Batch)) %>%
  rename(BatchLiverLipid = Batch)
rownames(covar_small) <- as.numeric(str_replace(covar_small$Mouse.ID, "DO-", ""))
covar_small$Mouse.ID <- NULL
```

```{r}
covar_small$BatchPlasmaLipid <- as.character(plasmaz$Batch)
covar_small$BatchLiverMetab <- str_replace(metabz[matchM2, "Batch"], "Batch", "")
tmp <- rep(NA, nrow(covar_small))
tmp[matchC2] <- as.character(cecumz$Batch)
covar_small$BatchCecumLipid <- tmp
```

```{r}
saveRDS(covar_small, file.path(datapath, "molecule", "covar_small.rds"))
```


#### analyses_tbl info

**Now need to create analyses_tbl object here.**
**First have to get peaks**

#### More detailed look at liver metab

```{r}
library(ggplot2)
metabn <- readRDS(file.path(datapath, "molecule",
                            "attie_liver_metabolites_zscore_uwisc_normalized.rds"))
```

```{r}
dat <- data.frame(x = unlist(metabn[104,-(1:13)]), 
                  y = unlist(metabz[104,-(1:13)]))
```

```{r}
ggplot(dat, aes(x,y)) + geom_point()
```

```{r}
dat <- data.frame(x = unlist(metabn[161,-(1:13)]), 
                  y = unlist(metabz[106,-(1:13)]))
```

```{r}
ggplot(dat, aes(x,y)) + geom_point()
```

```{r}
dat <- data.frame(x = unlist(metabn[106,-(1:13)]), 
                  y = unlist(metabz[106,-(1:13)]))
```

```{r}
ggplot(dat, aes(x,y)) + geom_point()
```


```{r}
dat <- data.frame(x = unlist(metabn[161,-(1:13)]), 
                  y = unlist(metabz[161,-(1:13)]))
```

```{r}
ggplot(dat, aes(x,y)) + geom_point()
```
