---
title: "Small Molecules"
author: "Brian S. Yandell"
date: "8/29/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

See Box folder [MS data sets for DO project](https://uwmadison.app.box.com/folder/33549545179)
and subfolders

* RDS files for all QTLs
* Gatti R scripts
* Raw and UW normalized data files

See also [Dan Gatti's AttieMetabolomics GitHub](https://github.com/dmgatti/AttieMetabolomics) and
[Dan Gatti's FTP site](ftp://ftp.jax.org/dgatti/_for_Mark/liver/).

```{r}
datapath <- "~/Documents/Research/attie_alan/DO/data/DerivedData/"
```

```{r}
livern <- readRDS(file.path(datapath, "jax", "attie_liver_lipids_normalized.rds"))
```

```{r}
liverz <- readRDS(file.path(datapath, "jax", "attie_liver_lipids_zscore_normalized.rds"))
```

The following is the scan1 object.

```{r}
liver <- readRDS(file.path(datapath, "jax", "liver_lipids_jax_norm_all_qtl.rds"))
```

```{r}
liverd <- readRDS(file.path(datapath, "jax", "liver_lipid_pheno.descr.rds"))
```

The `pheno.descr` file is different from what Karl used. Further, some of the phenotypes are mixed in with the phenotype table, and the `Mouse.ID` may be somewhat different. All fixable but will take careful examination. 
